---
title: "Model ThreeMe Tunisia"
author: "Lucia SEQUEIRA - Bijan VALILOU - Jin WANG"
date: "January 2022"
output:
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
  pdf_document:
    citation_package: biblatex
    toc: yes
    number_sections: yes
    toc_depth: 2
    highlight: tango
    latex_engine: lualatex
    keep_tex: yes
subtitle: 'Macroeconomic impacts of National Low-Carbon Strategy for a little country in development : the case of Tunisia'
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyfoot[CO,CE]{Modèle ThreeMe Tunisie}
- \fancyfoot[LE,RO]{\thepage}
bibliography: references.bib
---

```{r, echo = FALSE, tidy = TRUE, include = FALSE}

# Importer des packages
library(knitr) 
library(rmarkdown) 
library(markdown) 
library(readxl)
library(tibble)
library(ggplot2)
library(dplyr)
library(usethis)

```

```{r, echo = FALSE, tidy = TRUE, include = FALSE}

# Importer de la base de données
sheet_names <- excel_sheets("Page Reporting_shock des scénarios.xlsx")
nb_sheets <- length(sheet_names)
for (i in 1:nb_sheets){  
        name <- sheet_names[i] 
        name2 <- paste0(name) 
        data <- read_excel("Page Reporting_shock des scénarios.xlsx", sheet = i) 
        data <- column_to_rownames(data, var = '_date_')
        colnames(data) <- c(2015:2050)
        data <- data.frame(t(data))
        data$year <- c(2015:2050)
        assign(name2,data)
        rm(name,name2,data)
} 

Sc1 <- (S1/Baseline-1)*100
Sc2 <- (S2/Baseline-1)*100
Sc3 <- (S3/Baseline-1)*100
Sc4 <- (S4/Baseline-1)*100
Sc5 <- (S5/Baseline-1)*100
Sc6 <- (S6/Baseline-1)*100
Sc1$year <- c(2015:2050)
Sc2$year <- c(2015:2050)
Sc3$year <- c(2015:2050)
Sc4$year <- c(2015:2050)
Sc5$year <- c(2015:2050)
Sc6$year <- c(2015:2050)

sheet_names <- excel_sheets("Page Macro.xlsx")
nb_sheets <- length(sheet_names)
for (i in 1:nb_sheets){  
        name <- sheet_names[i] 
        name2 <- paste0(name,"_Macro") 
        data <- read_excel("Page Macro.xlsx", sheet = i) 
        data <- column_to_rownames(data, var = '_date_')
        colnames(data) <- c(2015:2050)
        data <- data.frame(t(data))
        data$year <- c(2015:2050)
        assign(name2,data)
        rm(name,name2,data)
} 
```



\newpage

# Introduction

# Literature review

## Energy and economy framework in Tunisia (Jin)

## Climate policy

### Carbon tax (Jin)

### Energetic subsidy (Bee)

Subsidies are defined by @demoor1997 as *'any measure that keeps prices for consumers below the market level or keeps prices for producers above the market level or that reduces costs for consumers and producers by giving direct or indirect support'*. Energy subsidies are a common policy. Their amount is estimated at \$4.7 trillion in 2015 as pointed out @coady2019, which is equivalent to 6.3 percent of Gross Domestic Product . Energy subsidies are very high and fluctuate depending on the price of the energy products. From the database of the International Energy Agency, we can notice that the fossil fuel subsidies have [fallen by 42 percent between 2019 and 2020](https://www.iea.org/topics/energy-subsidies) due to the fall of fuel prices, but theses prices have rocketed in 2021.

The energy subsidies are very present in in the Middle East, North Africa, Afghanistan and Pakistan region (MENAP). According to @coady2019, MENAP is the fourth region in absolute terms, which subsidized the most energy in 2015. Nevertheless, in relative terms, MENAP is the second, if we take the percent of its GDP. The prevalence of energy subsidies in MENAP can be explained by post-par period. @verme2017 explain that theses energy subsidies were introduced in Middle East and North Africa region (MENA) in order to stabilize prices after the decolonization. These stabilization mechanisms became a social protection system.

### Recycle of gouvernment income for tax (Lucia)

# Methodology

## ThreeME model (Bee)

### A neo-Keynesian Computable General Equilibrium model

### ThreeMe model Tunisia : Adaptation for a little country in development

## Description of scenario

-   describe redistribution

## Choice of evaluation indicators

### Data structure

All the indicators used in the analysis are in exact hat algebra, meaning the proportional variation to Baseline scenario.

### Kaya identity (Jin)

The Kaya identity, firstly proposed by \[Kaya 1990 a trouver\], is an identity where the total emission of carbon dioxide can be explained by four product driving forces as population, Gross Domestic Product (GDP) per capita, enerny intensity over GDP and carbon intensity over energy consumption [@kayaide2021]. It is expressed in the form:

$$ C = POP \times\frac{GDP}{POP} \times \frac{TEC}{GDP} \times \frac{C}{TEC} \tag{1}$$

where:

-   POP is global population;
-   GDP is gross domestic product;
-   TEC is total energy consumption;
-   C is total emission of carbon dioxide;

And:

-   GDP/POP is GDP per capita describing the economical activities within a period;
-   TEC/GDP is energy intensity;
-   C/TEC is carbon intensity;

In this study, we introduced Kaya identity to explain how different driving forces influenced the total emission for different scenarios. Firstly, a modified Kaya identity is used to analysis CO2 emission with the aggregated factors, then an extended Kaya identity coupled with Log Mean Divisia Index (LMDI) method to decomposite CO2 emission at the sectorial level. We will describe the detailed method in the following sections.

#### Aggregated level

We modified the function of Kaya identity mentioned above in order to adapt our model assumption, seeing the equation (2).

$$ C = VA \times\frac{TEC}{VA} \times \frac{C}{TEC} = E \times I \times P \tag{2}$$ 

where VA, TEC and C denote total added value of all economic sectors, total energy consumption and total CO2 emission. In the modified Kaya identity, we drew 3 driving forces for CO2 emission: E, I and P represent, respectively, overall economic activity, energy intensity over added value and carbon intensity over energy consumption. Here we cancelled population effect because its increasing rate remains still over time for the 6 scenarios and is considered as an exogenous variable in ThreeME model.

#### Sectorial level

# Results et discussions

## Comparasion of redistribution (lucia)

-   graphique PIB : grandes tendances des scénarii Taxe carbone, Levée des subventions énergétiques
-   chomage : grandes tendances des scénarii Taxe carbone, Levée des subventions énergétiques
-   émissions : grandes tendances des scénarii Taxe carbone, Levée des subventions énergétiques

## Carbon tax (jin)

Avec/ sans recyclage Focus inter-sectoriel

## Lifting of energy subsidies (bee)

Avec/ sans recyclage Focus inter-sectoriel

## ENR

## SNBC

# Prolongements

## D'autres leviers pour analyser (ouverture)

## L'amélioration du cadre statistique

OUverture : pb du secteur informel non pris en compte par la comptabilité nationale

# Conclusion

# Bibliographie
